\section{Introduction}

\subsection{General Background}

Geological media comprise natural solid Earth materials such as sedimentary, igneous, and metamorphic rocks, together with the fractured, weathered, porous, and stress-bearing rock masses derived from them. From a geological perspective, these materials are products of deposition, diagenesis, magmatism, metamorphism, tectonic deformation, and surface alteration; from a physical perspective, they are solids with complex internal architecture and variable response to external disturbance. Their behavior depends on mineral composition, density, porosity, pore structure, grain contacts, elastic stiffness, and the thermal and stress conditions under which they exist. Rock physics is concerned with the relation between geophysical observations and these underlying material properties, whereas geology explains why those properties vary from one rock body to another. For this reason, a geological medium cannot be treated adequately as an abstract uniform solid alone: its response reflects both its present physical state and its geological history \cite{mavko2009,jaeger2007,robertson1988,turcotte2014}.

This dependence on internal structure is fundamental because even rocks belonging to the same broad lithological group may behave quite differently under different crustal conditions. Thermal properties vary with composition, porosity, water content, pressure, and temperature, while elastic response and seismic velocity vary with lithology, packing, fracture density, consolidation, and effective stress. In porous materials, grain-to-grain contact and pore geometry influence how efficiently stresses are transmitted and how efficiently heat is conducted; in fractured materials, microcracks, joints, and bedding discontinuities alter both stiffness and continuity. Consequently, the relevant properties of a geological medium are not single fixed constants, but effective characteristics emerging from composition, microstructure, and in situ conditions. This is particularly important in rocks, where heterogeneity occurs over many scales, from mineral grains and pores to fractures, layers, and larger structural domains \cite{mavko2009,jaeger2007,wolff1982}.

\subsection{Physical and Geological Motivation}

Within geophysics, the propagation of waves through rocks is one of the principal means by which the subsurface becomes observable. Seismic methods use elastic disturbances traveling through geological media to infer layer boundaries, depth to bedrock, stratigraphic contrasts, lateral heterogeneity, and the geometry of buried structures. In engineering geology and near-surface investigation, wave-based methods help bridge the gaps between boreholes and excavations by providing laterally continuous information about the stiffness and stratification of the ground. The practical value of wave propagation therefore extends beyond theoretical mechanics: it underpins seismic interpretation, site characterization, geotechnical assessment, and broader subsurface characterization in both academic and applied contexts \cite{aki2002,lucius2007,mavko2009}.

The importance of this subject becomes even clearer when one considers geological hazards and energy resources. Near-surface shear-wave velocity is a major factor in the amplification and duration of earthquake ground motion, which means that the elastic properties of rocks and sediments directly affect seismic hazard assessment and engineering response. At the same time, geothermal systems provide a strong physical motivation for studying thermally affected rock behavior. The exploitation of geothermal energy depends not only on the presence of heat in the subsurface, but also on how that heat is transferred through rocks and how the rock framework responds mechanically to changing temperature conditions. More broadly, thermal properties of rocks are essential for heat-flow studies, geothermal resource appraisal, and other crustal thermal problems. Thus, the study of wave propagation in thermally influenced geological media is relevant to geophysics, engineering geology, geothermal systems, and geological hazard analysis alike \cite{turcotte2014,robertson1988,jaeger2007}.

\subsection{Thermal--Mechanical Interaction in Rocks}

This practical significance leads directly to the central physical issue of the present work: the interaction between temperature and deformation. In classical thermoelasticity, a change in temperature produces thermal expansion or contraction. If a body were completely free to deform, this response would appear as thermal strain alone. Geological media, however, are rarely free in this sense. Rock layers are confined by neighboring strata, mineral grains are restrained by adjacent grains, and rock masses are bounded by joints, faults, lithological contacts, and pre-existing stress fields. Under such conditions, thermal expansion or contraction may generate thermal stresses. Classical thermoelastic theory treats these processes as the coupled interaction of temperature, strain, and stress, and describes them through the combined framework of elasticity and heat conduction \cite{nowacki1986,boley1960}.

The geological importance of thermal stress is not merely theoretical. Temperature variation can contribute to joint formation, cracking, and progressive damage in rocks. In igneous settings, cooling and contraction are recognized causes of jointing, including columnar patterns in basalt and related fracture sets in other massive rocks. At smaller scales, differences in mineral expansion and contraction may generate local stress concentrations and microcracks within polymineralic aggregates. Repeated thermal cycling can therefore modify stiffness, fracture density, and long-term rock integrity. In this sense, temperature is not just an external background variable added to a mechanical problem; it may actively reshape the physical state of the medium through which later disturbances propagate \cite{turcotte2014,jaeger2007,robertson1988}.

\subsection{Heat Conduction and Elastic Wave Propagation}

A further point must be emphasized before wave motion itself is considered. Thermal disturbances and mechanical disturbances do not spread through rocks by the same physical mechanism. In the classical framework, temperature evolves primarily through heat conduction. In rocks, conduction is the principal mechanism of heat transport in the solid framework, and the rate of thermal adjustment depends on thermal conductivity and heat capacity, which together define thermal diffusivity. A material with high thermal diffusivity responds more rapidly to a sudden external temperature change than one with low diffusivity. Mechanical disturbances, by contrast, propagate as elastic waves governed chiefly by density and elastic moduli. Thermoelasticity links these processes by recognizing that temperature changes generate stress and strain, while deformation may influence the thermal state of the body. The resulting problem is therefore one of coupled evolution rather than two independent fields developing side by side \cite{nowacki1986,robertson1988}.

The elastic-wave aspect of this coupled problem is introduced most naturally through body waves in solids. At the conceptual level, P-waves are compressional disturbances associated with local volume change, whereas S-waves are shear disturbances associated mainly with change of shape. In a given material, P-waves travel faster than S-waves, and the velocities of both depend on density and elastic stiffness. When such waves encounter contrasts in density or elastic properties, part of the energy is transmitted and part is reflected, which is why wave observations can be used to infer subsurface structure. This classical picture is essential because it provides the physical basis for interpreting how geological media respond dynamically to transient loading and how material contrasts reveal themselves through wave behavior \cite{aki2002,mavko2009}.

\subsection{Directional Behavior in Geological Wave Propagation}

For the present thesis, it is especially important to make the notion of directional behavior explicit. In an ideal isotropic body, wave speed is independent of direction. Geological media rarely satisfy this idealization. Bedding in sedimentary sequences, foliation in metamorphic rocks, preferred mineral orientation in crystalline rocks, aligned pores and cracks, and stress-induced fracture closure or opening can produce anisotropy. As a result, waves traveling parallel to a structural fabric may propagate with different speed, attenuation, or particle-motion characteristics than waves traveling across it. Seismic anisotropy therefore expresses the internal ordering of the medium and helps link observed wave behavior to geological structure \cite{aki2002,crampin1981,thomsen1986}.

Directional behavior is not confined to elasticity alone. Thermal conductivity may also be anisotropic in layered or foliated materials, so the temperature field itself can evolve differently along and across the rock fabric. In thermoelastic problems, the coupled response may therefore inherit direction dependence from both the elastic and thermal properties of the medium. This means that a thermal disturbance and the related mechanical response may not spread uniformly in all directions, especially when the rock contains bedding, foliation, oriented cracks, fracture sets, or preferred mineral fabrics. Directional prediction is therefore physically meaningful because it attempts to estimate how the response changes with propagation direction in a structured geological medium.

\subsection{Research Gap}

Against this background, an important gap can be identified. The component theories needed to understand the present problem are well established, but they are often presented in separate disciplinary frameworks. Rock physics emphasizes the relation between geophysical observables and rock properties; quantitative seismology develops the theory of elastic-wave propagation; classical thermoelasticity develops the coupling between deformation and temperature; and rock mechanics and geodynamics discuss thermal stress and fracture in geological settings. What is less commonly offered at an introductory level is a single physical-geological synthesis that brings these elements together specifically for thermoelastic wave propagation in rocks, while also preserving the directional implications of geological fabric and heterogeneity \cite{mavko2009,aki2002,nowacki1986,boley1960,jaeger2007}.

A second gap concerns the scope of prediction in this project. Thermoelastic wave propagation can be studied through direct numerical solutions of governing equations, for example by discretizing the equations of elasticity and heat conduction. Such approaches are important in computational mechanics and geophysics, and they may be discussed separately as background or related work. However, the present thesis does not aim to develop a new numerical solver or to reproduce full numerical simulations of thermoelastic wave propagation by classical numerical methods. Instead, the project is positioned around prediction: the physical theory is used to define the relevant variables, material parameters, and directional meaning of the problem, while the practical part focuses on estimating the directional behavior of thermoelastic wave propagation using predictive models. Thus, the gap addressed here is not the absence of numerical methods in the field, but the need for a physically grounded predictive formulation that remains consistent with geological media and thermoelastic theory.

\subsection{Problem Statement}

The physical problem considered in this thesis may therefore be formulated in classical terms as follows. Given a geological material with density, elastic properties, coefficient of thermal expansion, thermal conductivity, and heat capacity, one needs to understand how a thermal disturbance and the associated mechanical deformation are related within the medium. More specifically, the relevant quantities are the temperature field, displacement, strain, and stress, together with their dependence on material properties and propagation direction. Temperature changes generate thermal strain; constrained thermal strain gives rise to stress; and these stresses interact with elastic-wave propagation. In geological media, this problem must always be interpreted with awareness of heterogeneity, anisotropy, and the influence of in situ conditions, even when the theoretical starting point is classical continuum theory \cite{nowacki1986,robertson1988,aki2002,mavko2009}.

In this thesis, the problem is not treated as a task of obtaining full-field numerical solutions of the governing equations. The role of the theoretical formulation is to clarify what physical quantities are involved, why the direction of propagation matters, and which rock properties control the response. On this basis, the practical task is understood as directional prediction rather than direct numerical simulation. This distinction is important: numerical methods may serve as established background in the broader literature, but the present work uses the physical and geological framework to support model-based prediction of thermoelastic wave propagation behavior in geological media.

\subsection{Aim and Scope of the Theoretical Part}

The aim of the theoretical part of this thesis is to establish the physical and geological basis necessary for understanding directional prediction of thermoelastic wave propagation in rocks. To achieve this aim, the discussion identifies the principal properties of geological media that control thermal and mechanical response, reviews the classical description of elasticity and elastic-wave motion, explains heat conduction and thermal expansion, and clarifies how constrained temperature change produces thermoelastic stress. Particular attention is given to the fact that real geological media are structured materials in which both wave propagation and thermal evolution may exhibit directional dependence because of layering, foliation, fracture systems, and mineral fabric.

Accordingly, the Introduction and Chapter 1 remain within the scope of classical physical theory and geological interpretation. Their focus is limited to geological material properties, elasticity, thermal expansion, heat conduction, thermoelastic stress, elastic wave propagation, and direction-dependent behavior. They do not present a new finite-difference, finite-element, or other classical numerical solution of the governing thermoelastic equations. Later chapters may refer to existing numerical approaches as related work or as a source of physical context, but the purpose of the present Introduction is to establish the concepts required to understand why model-based directional prediction is physically meaningful for thermoelastic wave propagation in geological media.
