\clearpage

\renewcommand{\refname}{REFERENCES}

\begin{thebibliography}{99}

\bibitem{aki2002}
Aki, K., and Richards, P. G.,
\textit{Quantitative Seismology}, 2nd ed.
Sausalito, CA: University Science Books, 2002.

\bibitem{boley1960}
Boley, B. A., and Weiner, J. H.,
\textit{Theory of Thermal Stresses}.
New York: John Wiley \& Sons, 1960.

\bibitem{cornet2007}
Cornet, F. H.,
\textit{Elements of Crustal Geomechanics}.
Cambridge: Cambridge University Press, 2007.

\bibitem{jaeger2007}
Jaeger, J. C., Cook, N. G. W., and Zimmerman, R. W.,
\textit{Fundamentals of Rock Mechanics}, 4th ed.
Malden, MA: Blackwell Publishing, 2007.

\bibitem{mavko2009}
Mavko, G., Mukerji, T., and Dvorkin, J.,
\textit{The Rock Physics Handbook: Tools for Seismic Analysis of Porous Media}, 2nd ed.
Cambridge: Cambridge University Press, 2009.

\bibitem{nowacki1986}
Nowacki, W.,
\textit{Thermoelasticity}, 2nd ed.
Oxford: Pergamon Press; Warsaw: PWN--Polish Scientific Publishers, 1986.

\bibitem{robertson1988}
Robertson, E. C.,
\textit{Thermal Properties of Rocks}.
U.S. Geological Survey Open-File Report 88--441, 1988.

\bibitem{turcotte2014}
Turcotte, D. L., and Schubert, G.,
\textit{Geodynamics}, 3rd ed.
Cambridge: Cambridge University Press, 2014.

\bibitem{wolff1982}
Wolff, R. G.,
\textit{Physical Properties of Rocks: Porosity, Permeability, Distribution Coefficients, and Dispersivity}.
U.S. Geological Survey Open-File Report 82--166, 1982.

\bibitem{crampin1981}
Crampin, S.,
``A review of wave motion in anisotropic and cracked elastic-media,''
\textit{Wave Motion}, vol. 3, no. 4, pp. 343--391, 1981.

\bibitem{lucius2007}
Lucius, J. E., Langer, W. H., and Ellefsen, K. J.,
\textit{An Introduction to Using Surface Geophysics to Characterize Sand and Gravel Deposits}.
U.S. Geological Survey Circular 1310, 2007.

\bibitem{robertson1988}
Robertson, E. C.,
\textit{Thermal Properties of Rocks}.
U.S. Geological Survey Open-File Report 88--441, 1988.

\textit{Fundamentals of Rock Mechanics}, 4th ed.
Malden, MA: Blackwell Publishing, 2007.

\bibitem{mavko2009}
Mavko, G., Mukerji, T., and Dvorkin, J.,
\textit{The Rock Physics Handbook: Tools for Seismic Analysis of Porous Media}, 2nd ed.
Cambridge: Cambridge University Press, 2009.

\bibitem{kazakhRockPropertiesHighPT}

\textit{Физические свойства минералов и горных пород при высоких давлениях и температурах.}

\bibitem{vaswani2017}
Vaswani, A., Shazeer, N., Parmar, N., Uszkoreit, J., Jones, L., Gomez, A. N., Kaiser, \L., and Polosukhin, I.,
``Attention is all you need,''
\textit{Advances in Neural Information Processing Systems (NeurIPS)}, vol. 30, pp. 5998--6008, 2017.

\bibitem{liNeuralOperator2023}
Li, Z., Meidani, K., and Farimani, A. B.,
``Transformer for partial differential equations' operator learning,''
\textit{Transactions on Machine Learning Research}, 2023.

\bibitem{caoChooseTransformer2021}
Cao, S.,
``Choose a transformer: Fourier or Galerkin,''
\textit{Advances in Neural Information Processing Systems (NeurIPS)}, vol. 34, pp. 24924--24940, 2021.

\bibitem{wuTransolver2024}
Wu, H., Luo, H., Wang, H., Wang, J., and Long, M.,
``Transolver: A fast transformer solver for PDEs on general geometries,''
\textit{Proceedings of the 41st International Conference on Machine Learning (ICML)}, 2024.

```
\bibitem{openstax-waves}
William Moebs, Samuel J. Ling, Jeff Sanny. \textit{University Physics, Volume 1 — Chapter 16 “Waves”}.  
OpenStax, 2016.  
Online: \url{https://assets.openstax.org/oscms-prodcms/media/documents/UniversityPhysicsVol1-WEB.pdf} \cite{openstax-waves}

\bibitem{feynman}
R. P. Feynman, R. B. Leighton, M. Sands. \textit{The Feynman Lectures on Physics, Vol. I}, Ch. 51 “Waves”.  
California Institute of Technology.  
Online: \url{https://www.feynmanlectures.caltech.edu/I_51.html} \cite{feynman}

\bibitem{mit803}
MIT OpenCourseWare. \textit{8.03SC Physics III: Vibrations and Waves}.  
Lections + notes. 2016.  
Online: \url{https://ocw.mit.edu/courses/8-03sc-physics-iii-vibrations-and-waves-fall-2016/} \cite{mit803}

\bibitem{purdue-waves}
Purdue University, Lecture Notes on Waves and Oscillations.  
Available on cite Physics Department Purdue \url{https://www.physics.purdue.edu/} \cite{purdue-waves}

\bibitem{ispulov2022}
Ispulov, N. A., Zhumabekov, A. Zh., Qadir, A.,
Kurmanov, A. A., Sarymova, Sh. N., Dossumbekov, K. R.,
and Arinov, E.,
\href{https://pdfs.semanticscholar.org/488a/91fa54fe48b2a361f9a4caafd9e606447f7f.pdf}
{\textit{The Propagation of Thermoelastic Waves in Different Anisotropic Media Using Matricant Method}}.
\textit{Advances in Mathematical Physics}, vol. 2022,
Article ID 5787899, 8 pages, 2022.
\href{https://doi.org/10.1155/2022/5787899}
{doi:10.1155/2022/5787899}.

\bibitem{ismail2025}
Ismail, M. F., Ahmed, H. M., Safwat, S.,
Ramadan, M. E., Abdalla, N. S. E.,
and Soliman, M.,
\href{https://www.nature.com/articles/s41598-025-09344-w.pdf}
{\textit{Analytical Wave Solutions in Thermoelastic Media with Temperature-Dependent Properties via IMETF Method}}.
\textit{Scientific Reports}, vol. 15, Article no. 32553, 2025.
\href{https://doi.org/10.1038/s41598-025-09344-w}
{doi:10.1038/s41598-025-09344-w}.

\bibitem{hou2023}
Hou, W., Fu, L.-Y., and Carcione, J. M.,
\href{https://ricerca.ogs.it/retrieve/a16968a5-6f6f-496e-819f-d244ff797fbd/hou-et-al-2023-amplitude-variation-with-offset-in-thermoelastic-media.pdf}
{\textit{Amplitude-Variation-with-Offset in Thermoelastic Media}}.
\textit{Geophysics}, vol. 88, no. 1, pp. MR25--MR33, 2023.
\href{https://doi.org/10.1190/geo2021-0815.1}
{doi:10.1190/geo2021-0815.1}.

\bibitem{alipova2025}
Alipova, B.,
\href{https://ceur-ws.org/Vol-3966/W1Paper2.pdf}
{\textit{Distribution of Coupled Thermoelastic Waves in the Different Rocks Using MatLab GUI Application}}.
In: \textit{Proceedings of DTESI 2024: 9th International Conference on Digital Technologies in Education, Science and Industry},
CEUR Workshop Proceedings, vol. 3966, 2025.

\bibitem{yang2023}
Yang, B., Li, Z., Zeng, L., and Chen, X.,
\href{https://academic.oup.com/gji/article/232/2/1408/6761682}
{\textit{Simulation of Thermoelastic Wave Propagation in 3-D Multilayered Half-Space Media}}.
\textit{Geophysical Journal International}, vol. 232, no. 2,
pp. 1408--1426, 2023.
\href{https://doi.org/10.1093/gji/ggac401}
{doi:10.1093/gji/ggac401}.

\end{thebibliography}