\renewcommand* {\refname}{}
\begin{thebibliography}{}
\bibitem{b1} 
What diseases are insured against Kazakhstanis \url{https://kz.kursiv.media/2021-06-01/ot-kakikh-bolezney-strakhuyutsya-kazakhstancy/}
\bibitem{b2}What is cardiovascular disease? List of diseases, statistics and recommendations for the prevention of cardiovascular diseases. \url{https://profilaktica.ru/for-population/profilaktika-zabolevaniy/serdechno-sosudistye-zabolevaniya/chto-takoe-serdechno-sosudistye-zabolevaniya/}

\bibitem{b3} Atherosclerosis statistics
 \url{ http://surgeryzone.net/info/obshhemedicinskaya-informaciya/statistika-ateroskleroza.html}
\bibitem{b4}What is cardiovascular disease? List of diseases, statistics and recommendations for the prevention of cardiovascular diseases. \url{ https://profilaktica.ru/for-population/profilaktika-zabolevaniy/serdechno-sosudistye-zabolevaniya/chto-takoe-serdechno-sosudistye-zabolevaniya/ }
\bibitem{b5} Types of diseases of the cardiovascular system
\url{https://chekhovsc.ru/blog/vidy-zabolevanij-serdechno-sosudistoj-sistemy }
\bibitem{b6}
Hydrodynamics: studies. Manual for students of non-mathematical faculties / A.B. Mazo, K.A. Potashev. – Kazan: Kazan. un-t, 2013. – 2nd ed. – 128 p.
\bibitem{b7}
J.H. Forrester and D.F. Young, Flow through a converging-diverging tube and its implications in occlusive vascular stenosis-I, Theoretical development, J. Biomech. 3, 297-316, (1970).
\bibitem{b8}O. S. Petrova, N. Y. Grishina "Computer modeling of cardiac activity"\url{ https://pandia.ru/text/79/027/9026.php}
\bibitem{b9} A. N. Tikhonov and A. A. Samarskii, Equations of mathematical physics. Courier Corporation, 2004.
\bibitem{b10} Mathematical models of hydrodinamics \url{ http://math.phys.msu.ru/archive/2014_2015/51/12_MMG_Davidova.pdf}
\bibitem{b11} Zaitsev A.V. "Numerical Research of MAthematical Model of the Current in the Channel"
\url{https://cyberleninka.ru/article/n/chislennoe-issledovanie-matematicheskoy-modeli-techeniya-v-kanale-1/viewer}
\bibitem{b12} © 2012 Kiselev I.N., Semisalov B.V., Biberdorf E.A.3
, Sharipov R.N., Blokhin A.M.3, Kolpakov F.A.,"Modular modeling of human cardiovascular systems"
\url{https://www.matbio.org/2012/Kiselev_7_703.pdf}
\bibitem{b13}Vinicius C. Rispoli,corresponding author Jon F. Nielsen, Krishna S. Nayak, and Joao L. A. Carvalho "Computational fluid dynamics simulations of blood flow regularized by 3D phase contrast MRI"
\url{https://www.ncbi.nlm.nih.gov/pmc/articles/PMC4661988/}
\bibitem{b14} Cardiovascular system \url{ https://www.tromboza.net/vsyo-o-tromboze/cardiovascular-system}
\bibitem{b15} J.C. Misra and B.K. Sahu, Flow through blood vessels under the action of a periodic acceleration field: A mathematical analysis, Computers Math. Applic. 16 (12), 993-1016, (1988).
\bibitem{b16}  European Journal of Cardio-thoracic Surgery 33 (2008) 164—167 

\bibitem{b17}E. Belardinelli, M. Ursino and E. Lemmi, A preliminary theoretical study of arterial pressure perturbations under shock accelerations, ASME J. Biomech. Engng. 111, 233-240, (1989).
\bibitem{b18} G.R. Cokelet, The rheology of human blood, In Biomechanics-Its Foundations and Objectives, (Edited by Y.C. Fung, N. Perrone and M. Anliker), pp. 63-103, PrenticeHall, (1981).
\bibitem{b19} T. Murata, Theory of non-Newtonian viscosity of red blood cell suspension: Effect of red cell deformation, Biorheol. 20, 471-483, (1983).
\bibitem{b20} Jeremie Labroquere Navier-Stokes computation with num3sis – Poiseuille flow and Backward facing step
\bibitem{b21} Issakhov A., Mathematical modeling of the discharged heat water effect on the aquatic environment from thermal power plant // International Journal of Nonlinear Science and Numerical Simulation, – 2015, 16(5), pp. 229–238, doi:10.1515/ijnsns-2015-0047
\bibitem{b22} 7 th International Conference on Computational Heat and Mass Transfer , CFD analysis of blood flow in human aorta with experimental validation.
\bibitem{b23} Journal of Biomechanical Engineering Copyright © 2005 by ASME october 2005,Vol 127 , 3-D Numerical Simulation of Blood Flow Through Models of the Human Aorta
\bibitem{b24}  Laura McKinney "The difference between compressible and incompressible fluids" \url{https://ru.strephonsays.com/difference-between-compressible-and-incompressible-fluids}
\bibitem{b25} Bretscher M. S. The molecules of the cell membrane. Scientific American. 1985.253(4)
\bibitem{b26} International Journal of Experimental Education. - 2015. – No. 12-4. – pp. 539-543;
\bibitem{b27} The Gauss — Seidel method \url{ https://dic.academic.ru/dic.nsf/ruwiki/408798}
\bibitem{b28} Zhazykpaev A.K. Torshina O.A. "Numerical method for solving a mixed boundary value problem generated by the Laplace differential equation"\url{https://eduherald.ru/ru/article/view?id=19395}
\bibitem{b29}Finite time blowup for an averaged three-dimensional Navier-Stockes Equation\url{https://arxiv.org/pdf/1402.0290v2.pdf}
\bibitem{b30} Kazakh Professor solved one of the most difficult mathematical tasks of the millennium \url{ https://www.math.kz/images/journal/2013-4/Otelbaev_N-S_21_12_2013.pdf}
\bibitem{b31} The blood system. Composition, quantity, properties, basic functions of blood. Basic physiological constants of blood \url{https://znzn.ru/normalnaya-fiziologiya/sistema-krovi}
\bibitem{b32}J. F. Thompson, B. K. Soni and N. P. Weatherill. Handbook of Grid Generation. CRC Press, Boca Raton, 1999.
\bibitem{b33} A.A. Degtyarev, Finite difference method, Samara 2011.
\bibitem{b34} The programming language - Python \url{https://blog.skillfactory.ru/glossary/python/}
\bibitem{b35} The numpy library \url{https://python-school.ru/wiki/numpy/}
\bibitem{b36} The matplotlib library \url{https://skillbox.ru/media/code/biblioteka-matplotlib-dlya-postroeniya-grafikov/}
\bibitem{b37} The SPH method \url{https://www.dive-solutions.de/blog/sph-basics}
\bibitem{b38} Pulsatile flow \url{https://en.wikipedia.org/wiki/Pulsatile_flow}
\bibitem{b39} The Reynolds number \url{https://elementy.ru/trefil/21200/Chislo_Reynoldsa}
\bibitem{b40} The history of Python \url{https://techrocks.ru/2019/01/21/about-python-briefly/}
\bibitem{b41}A. R. Mitchell and D. F. Griffiths, “The finite difference method in partial differential equations,” A Wiley-Interscience Publication,1980.
\bibitem{b42} N. Vatin, A. Girgidov, and K. Strelets, “Numerical modelling the three-dimensional velocity field in the cyclone,” Magazine of Civil Engineering, 2011.

\end{thebibliography}