\subsection{Model Comparison Framework}

The methodology uses a common physical input-output structure to compare several predictive model families. The comparison is not intended to prove that one model fully solves the thermoelastic wave propagation problem. Instead, the models are treated as alternative predictive components that operate within the same physically motivated framework. This makes it possible to compare their behavior using consistent material parameters, thermal conditions, source--probe geometry, and output interpretation.

The first model family is the Physics-Informed Neural Network, or PINN. In this thesis, the PINN is treated as a physics-informed baseline because it is the only compared component whose training objective explicitly includes thermoelastic residuals derived from the governing equations introduced in Chapter 2. Its role in the comparison is not to claim a complete solution of the full three-dimensional coupled thermoelastic system, but to provide a reference model whose structure is directly guided by the strain--displacement relation, the thermoelastic constitutive law, the equation of motion, and the coupled heat equation.

The second model family is the Fourier Neural Operator, or FNO. Operator-learning methods are relevant because thermoelastic wave propagation can be interpreted as a mapping from input fields, material parameters, and conditions to output response fields or response quantities. In this thesis, the FNO component is treated as a neural operator baseline trained on physically meaningful fields within the same input-output contract. However, in the current implementation it does not explicitly enforce the full thermoelastic partial differential equation system through physics residuals in the same way as the PINN.

The third model family is MeshGraphNet. Graph-based models are useful for problems where geometry, local neighborhoods, or mesh-like relationships are important. In the context of geological media, this is relevant because spatial structure and directional relationships influence wave behavior. Here, MeshGraphNet is treated as a graph-based predictive component rather than as a complete geological simulator. In the current prototype, its practical behavior may depend on the availability of checkpoint-backed model artifacts, and some runs may rely on a fallback response instead of a full learned rollout.

The fourth component is a Transformer-style baseline. This model family is included as an experimental supervised sequence/operator-style baseline for structured field prediction. Its role is to provide an additional point of comparison against the more physics-oriented or geometry-oriented approaches. As with the other model families, it is evaluated within the same physically motivated input-output framework, but in the current version it does not include explicit thermoelastic PDE residuals in the training objective.

These approaches are compared as alternative predictive components within the same physically motivated input-output framework. The main purpose of the comparison is to examine how different model families respond to the same material parameters, thermal conditions, and propagation directions. All models therefore share a common physical interpretation of inputs and outputs, but they use the physical formulation differently. PINN explicitly incorporates thermoelastic residuals derived from the governing equations, whereas FNO, MeshGraphNet, and the Transformer-style baseline remain data-driven or architecture-based predictive baselines that operate on physically meaningful variables without enforcing the full thermoelastic PDE system in the same way. Detailed software architecture, API behavior, service-level implementation, and deployment structure are not part of this chapter and are reserved for the implementation chapter.
