\subsection{Implementation and System Architecture}

This section describes the software implementation used to realize the prediction methodology introduced in Chapter~4. The purpose of the implementation is not to reproduce a full industrial geophysical simulator, but to provide a research prototype in which several predictive model families can be compared under one physically motivated workflow. In this sense, the system connects user-defined thermoelastic scenarios, effective geological material parameters, model routing, and normalized prediction summaries into one reproducible experimental stack. The implementation therefore acts as the practical bridge between the geological motivation of Chapter~1, the thermoelastic variables and governing relations of Chapter~2, and the comparative prediction task formulated in Chapter~3.

At the architectural level, the repository is organized as a service-oriented prototype. The main repository documentation describes a full local stack composed of a \texttt{FastAPI} backend orchestrator, a native \texttt{HTML/CSS/JavaScript} frontend, separate model services, a JSON-based geological medium catalog, and \texttt{Docker Compose} orchestration for local startup and testing. In practical terms, the system can be understood as a pipeline in which the user configures a geological medium and a thermoelastic scenario in the frontend, the backend validates and enriches the request, the selected model service performs prediction, and the normalized response is returned for display, comparison, and summary reporting. This design is appropriate for the thesis because the main scientific question is comparative directional prediction rather than monolithic solver construction.

The frontend is the user-facing layer of the prototype. Its role is to provide a compact scientific interface for selecting a geological medium, choosing a prediction route, specifying the source and probe geometry, defining thermal and mechanical conditions, and visualizing the returned prediction summary. The repository documentation indicates that the frontend dynamically loads media and model information from the backend and presents a two-dimensional domain view for interactive experiments. From the perspective of the thesis, the frontend should therefore be described as an experimental configuration and visualization surface rather than as a source of physical modelling. It does not perform the thermoelastic computation itself. Instead, it exposes the structured inputs needed by the prediction workflow and displays the resulting directional quantities and field-level summary indicators.

The backend is the central orchestration layer of the system. This distinction is important: the backend is not the machine-learning model itself. According to the repository \texttt{README.md}, it validates a unified prediction request, resolves the geological medium from the catalog, merges scenario inputs with rock presets, routes the request to the selected remote model service, and normalizes the response before returning it to the caller. The backend schema in \texttt{backend/app/schemas/prediction.py} confirms this role. It enforces a common request structure containing the selected model, a \texttt{medium\_id}, scenario parameters, source coordinates and direction, probe coordinates, and domain description. It also validates geometric consistency, for example by requiring \texttt{nz = 1} and \texttt{lz = 0} when \texttt{domain.type = rect\_2d}. As a result, the backend provides the common contract that allows different model services to be compared under the same physical input description.

The geological medium catalog is stored in \texttt{backend/data/media/catalog.json} and functions as a compact repository of effective material presets. The catalog includes media such as sandstone, limestone, basalt, and granite together with representative properties such as density, total and effective porosity, \(V_p\), \(V_s\), thermal conductivity, heat capacity, and thermal expansion. These entries are not full geological databases and should not be interpreted as field-validated constitutive models. Instead, they serve the thesis purpose of supplying effective material parameters for locally homogeneous and isotropic thermoelastic modelling. This is fully consistent with the simplifications introduced in Chapters~1 and~2, where real geological media are recognized as heterogeneous, porous, fractured, layered, and often anisotropic, but the mathematical formulation proceeds through effective continuum parameters.

The repository separates the predictive components into distinct services. This separation is useful because it makes the comparison framework explicit and prevents the backend from being mistaken for a single model implementation. The \texttt{pinn-service} is the closest component to the governing equations of Chapter~2. Its documentation states that the service predicts the primary fields \((T,u,v,w)\) and that training includes supervised loss, velocity consistency, elastic wave residuals, and a coupled thermal residual. The architecture notes further state that the residual terms are built from the thermoelastic stress tensor, displacement gradients, and the coupled heat equation with the \(\gamma T_0 \, \partial \varepsilon_{kk}/\partial t\) term. For this reason, the PINN is the only explicitly physics-informed baseline in the current comparison.

The \texttt{fno-service} is implemented as a dedicated Fourier Neural Operator route and is explicitly documented as a trainable \texttt{FNO2d} baseline. Its current purpose is to learn the evolution of physically meaningful fields on a regular grid, with targets such as temperature and displacement components. However, the service documentation also states that its MVP objective is a supervised loss of the form
\begin{equation}
\mathcal{L}_{\mathrm{FNO}}
=
\operatorname{MSE}(\widehat{\mathbf{Y}},\mathbf{Y})
+ 0.01 \, \operatorname{relativeL2}(\widehat{\mathbf{Y}},\mathbf{Y}),
\end{equation}
without explicit thermoelastic PDE residual enforcement. It is therefore described in the thesis as a neural-operator baseline rather than as a physics-constrained solver. This distinction matters because the model still operates on physically meaningful fields, yet it does not incorporate the governing thermoelastic equations in the same way as the PINN.

The \texttt{mgn-service} provides the graph-based route. Its repository documentation shows that it can run a MeshGraphNet-style rollout when dataset and checkpoint artifacts are configured, but it can also return a deterministic fallback response when these artifacts are unavailable and fallback is enabled. This means that the service is best described as a graph-based predictive component within a prototype architecture. It is relevant to the thesis because directional propagation in geological media is naturally linked to geometry, local neighborhoods, and mesh-like spatial relationships. At the same time, its practical interpretation must remain careful: depending on the run configuration, the returned result may reflect either learned graph-based prediction or fallback logic intended to keep the system operational for demonstration and interface testing.

The repository also contains a \texttt{transformer-service}, documented as an autoregressive Transformer-style baseline parallel to the PINN service. Its architecture uses tokens carrying coordinates, current-state fields, and material properties, and predicts \((T,u,v,w)\) at target coordinates. The same documentation states that the current version uses supervised data loss only and does not yet include physics-informed residuals. It should therefore be described in the thesis as an experimental supervised sequence/operator-style baseline. A further practical caution is that the repository documentation is not completely uniform: the service is present in the repository structure and in \texttt{docker-compose.yml}, but the service-specific \texttt{README.md} still labels backend and compose integration as a next-iteration task. For this reason, Chapter~4 should present the Transformer-style component as an experimental route whose maturity is lower than that of the main orchestration pattern.

Taken together, these services form a common comparative architecture rather than four physically equivalent solvers. All of them receive inputs motivated by the same thermoelastic problem definition, but they use the physics differently. The PINN is the only component whose training objective explicitly contains thermoelastic residuals derived from the governing equations. The FNO, MeshGraphNet, and Transformer-style routes use physically meaningful inputs and outputs, but they do not enforce the full coupled thermoelastic PDE system in the same way. This wording is important for consistency with Chapter~3 and for avoiding overstatement in the thesis.

Although the mathematical formulation in Chapter~2 is written in general three-dimensional form, the practical implementation of the prototype is predominantly two-dimensional or direction-reduced. The backend schema explicitly supports \texttt{rect\_2d} and \texttt{rect\_3d}, but the current MVP path emphasizes \texttt{rect\_2d} cases with \texttt{z = 0}, \texttt{lz = 0}, and \texttt{nz = 1}. The FNO service documentation likewise states that the present baseline operates on \texttt{Z = 1} slices and intentionally postpones native \texttt{FNO3d}. In such settings, a source and a probe are usually selected in a planar cross-section, the vertical component of the direction is set to zero, and the elevation angle is effectively \(\psi = 0\). This does not contradict the theory. Rather, it means that the theory is kept general in three dimensions, while the implementation evaluates selected source--probe paths or two-dimensional cross-sections that are practical for the available data and prototype model services.

The deployment structure of the project reinforces this interpretation. The \texttt{docker-compose.yml} file defines a local multi-service environment in which the frontend, backend, \texttt{pinn-service}, \texttt{fno-service}, \texttt{mgn-service}, and \texttt{transformer-service} are started as separate containers, each with its own health or readiness endpoint. This design is valuable for a bachelor-level research prototype because it isolates component responsibilities, makes failure modes more visible, and allows individual services to be replaced or upgraded without rewriting the entire stack. The repository also includes service-specific scripts and documentation for data preparation, training, checkpoint management, validation, and local startup. These support reproducibility and experimentation, even though they do not by themselves establish scientific validation.

In summary, Chapter~4 implements the methodology of Chapter~3 through a backend-centered orchestration stack that connects user inputs, effective geological material presets, model routing, and normalized directional prediction outputs. The implementation should be described as a comparative research prototype with physics-informed and data-driven components, not as a final validated three-dimensional thermoelastic simulator. Its main strengths are unified routing, explicit material handling, service modularity, and transparent model comparison. Its main limitations are equally important: the prototype is 2D-first, medium presets remain starter values, different model routes use physics at different depths, and fallback-capable services must be interpreted carefully when reporting comparative results.
